---
title: "Toronto TTC Data Analysis"
author: Shirley Chen
date: format(Sys.time(), "%d %B %Y")
format: html
editor: visual
bibliography: bibliography.bib
---

```{r}
#| include: false
### Workspace Setup ###
library(tidyverse)
library(janitor)
library(opendatatoronto)
library(ggplot2)
library(knitr)
library(usethis)
library(gitcreds)
```

## Introduction

The Toronto Transit Commission (TTC) is the public transportation system that runs throughout the areas within Toronto, Ontario, Canada. It is one of the largest public transportation systems in North America, and consists of the transportation methods of subway, bus, and streetcar.

## Analysis

### Yearly Ridership analysis

From the graph below, we can see the trends of TTC Ridership over the course of 16 years from 2008 - 2023.

```{r}
#| echo: false
#### Read in the data ####
raw_ttc_data <-
  read_csv(
    file = 
      "TorontoMeasureData.csv",
    show_col_types = FALSE,
    skip = 0
  )

cleaned_ttc_data <-
  clean_names(raw_ttc_data)

# Rename value column
cleaned_ttc_data <- cleaned_ttc_data |>
  rename(number_of_riders = value,
         month = period)

# Remove the first column and rearrange columns
cleaned_ttc_data <- cleaned_ttc_data |>
  select(number_of_riders, month, year)

# Sum up monthly values by year
yearly_ttc_data_sum <- cleaned_ttc_data |>
  group_by(year) |>
  summarise(number_of_riders = sum(number_of_riders))

ggplot(yearly_ttc_data_sum, aes(
  x = factor(year), 
  y = number_of_riders, 
  fill = factor(year))) + 
  geom_bar(stat = "identity") + 
  labs(title = "TTC Ridership by Year, 2008-2023",
       x = "Year",
       y = "Total Ridership")
```

We can see that the highest amount of riders occurred in 2016, whereas the lowest number of riders occurred in 2021. The reason for the drop in number of riders during 2021 and in the years following could be attributed to various reasons, but there's a high chance that majority of the reasons are related to the 2020 COVID-19 pandemic.

After the pandemic, many job modes became remote or hybrid. The TTC was a main mode of transportation for a lot of professional, working individuals; since many jobs were no longer required in-person, there were a lot less rider numbers. Further, the pandemic caused panic about germs in public spaces, so it would make sense that less people would want to take public transportation in years following the pandemic.

### 2020 Ridership Analysis

To provide a deeper analysis, I will be presenting the data from the year 2020, which was the year of the COVID-19 global pandemic. At least in North America, the pandemic, without a doubt, affected one's everyday routine. Below is a graph presents the monthly ridership for the year 2020.

```{r}
#| echo: false
  
  covid_ttc_data <- cleaned_ttc_data |>
    filter(year == 2020)
  
  custom_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", 
                    "Aug", "Sep", "Oct", "Nov", "Dec")
  covid_ttc_data$month <- factor(covid_ttc_data$month, 
                                 levels = custom_order)
  
  ggplot(covid_ttc_data, aes(
    x = month,
    y = number_of_riders, 
    fill = month)) +
  geom_bar(stat = "identity") +
  labs(title = "TTC Ridership in 2020",
       x = "Month",
       y = "Number of Riders")
```

As we can see, the ridership numbers were highest during January, whereas the lowest numbers were in April. However, we can see the numbers start to significantly decline in March, which is when the COVID-19 virus was declared a pandemic.
